---
title: "Welcome"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readODS)
library(DT)
```

## Course description

The objective of this course is to show students how statistics is used in practice to answer a specific question, by introducing a series of important model-based approaches.

The students will learn to select and use appropriate statistical methodologies and acquire solid and practical skills by working-out examples on real-world data sets from various areas including medicine, genomics, ecology, and others.

All analyses will be conducted with the R software, possibly with interfacing to Python. No strong knwoledge neither of R or Python programming is required (only basic scripting).

## Schedule (tentative)

**Teachers** : Julien Chiquet (lecture + 1 PC), Geneviève Robin (2 PC)

**Course Evaluation**: 2 individual homework assignements + a final exam/project 

**Course Language**: French with all material in English

**Numerus closus**: 80 students

1. Statistical tests (x1)

    - Two-populations comparison
    - Power analysis
    - Multiple Testing

2. Regression models (x2)

    - Linear and Non Linear Regression models
    - Nonlinear regression models
    - Inference Diagnostic, Model comparison

3. Mixed effects models (x2)

    - Linear mixed effects models
    - Nonlinear mixed effects models
    - EM algorithm

4. Mixture models and model-based clustering (x2)

    - Gaussian mixture models for data clustering
    - Stochastic Block Models for graph clustering
    - (Variational) EM algorithm

5. Model-based Dimension Reduction (x2)

    - Multivariate Gaussian model
    - Probabilistic Gaussian PCA
    - Generalized mixed effect models

<!-- 

```{r, echo = FALSE}
readODS::read_ods(path = "schedule.ods") %>% 
  datatable(options = list(pageLength = 24, dom = "t", ordering = FALSE), rownames = FALSE)
```

-->

