---
title: "Mixture models"
subtitle: "Exercices"
---

## Preliminary

Only functions from `R`-base and stats (preloaded) are required plus packages from the **tidyverse** for data representation and manipulation.

```{r mixture-model-lab-config, message = FALSE}
library(tidyverse)
theme_set(theme_bw())
```

## Faithful data

The `faithful` data consist of the waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.


```{r mixture-faitful-load}
data("faithful")
faithful %>% 
  ggplot() + aes(x=waiting) + geom_histogram() + xlab("waiting (mm)")
```

Fit variants of a mixture of two Gaussian distributions to the faithful data.

 1. assuming different proportions, means and variances for the 2 distributions

 2. assuming same variances 
 
 3. assuming same means
 
 4. assuming same proportions

### Hint: skeleton of EM for 1D mixture of Gaussian

We follow what has been done during the course: we define the function that computes the value of the likelihood for each point in each component. It will be used to compute the posterior probabilities in the E-step.

```{r auxiliary function}
## likelihood (density) of each point in each component
## data (x) is fixed while theta is variable
dcomponents <- function(theta, x) {
    mapply(
      function(pi, mu, sigma) pi * dnorm(x, mu, sigma),
      theta$pi, theta$mu, theta$sigma,
      SIMPLIFY = TRUE
    )
}
```

Then, we define an EM where the function that computes the M step is given as an argument:

```{r general EM}
mixture_gaussian1D <- 
  function(x,      # a n-vector of data
           theta0, # starting values of vector paramter
           M_step    = M_step_general, # an R function
           max_iter  = 100,
           threshold = 1e-6) {

  ## initialization
  n <- length(x)
  likelihoods  <- dcomponents(theta0, x) 
  deviance     <- numeric(max_iter)
  deviance[1]  <- -2 * sum(log(rowSums(likelihoods)))

  for (t in 1:max_iter) {
    
    # E step
    tau <- likelihoods / rowSums(likelihoods)

    # M step
    theta <- M_step(tau, x)
    
    ## Assessing convergence
    likelihoods   <- dcomponents(theta, x)
    deviance[t+1] <- - 2 * sum(log(rowSums(likelihoods)))

    ## prepare next iterations
    if (abs(deviance[t + 1] - deviance[t]) < threshold)
      break
    
  }

  list(theta = theta, deviance = deviance[t + 1])
}
```

## Epilepsy data

The data `seizures.csv` consists of  daily counts of epileptic seizures for 6 patients.

```{r mixture-seizures-load, message = FALSE}
seizures <- readr::read_csv('../../data/seizures.csv')
seizures %>% 
  ggplot() + aes(x = time, y = nsz) + geom_point() + facet_wrap( ~ id)
```
 
  1. Select the id 12 and fit a Poisson distribution to the number of seizures for this patient.

  2. Implement and use a EM algorithm for fitting a mixture of 2 Poisson distributions to this data
  
  3. Compare the two models
  
