---
title: "Mixture Models"
subtitle: "Exercices"
---

## Preliminary

Only functions from `R`-base and stats (preloaded) are required plus packages from the **tidyverse** for data representation and manipulation.

```{r mixture-model-lab-config, message = FALSE}
library(tidyverse)
theme_set(theme_bw())
```

## Faithful data

The `faithful` data consist of the waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.


```{r mixture-faitful-load}
data("faithful")
faithful %>% 
  ggplot() + aes(x=waiting) + geom_histogram() + xlab("waiting (mm)")
```

Fit variants of a mixture of two Gaussian distributions to the faithful data.

 1. assuming different proportions, means and variances for the 2 distributions

 2. assuming same variances 
 
 3. assuming same means
 
 4. assuming same proportions
 
## Epilepsy data

The data `seizures.csv` consists of  daily counts of epileptic seizures for 6 patients.

```{r mixture-seizures-load, message = FALSE}
seizures <- readr::read_csv('../../data/seizures.csv')
seizures %>% 
  ggplot() + aes(x = time, y = nsz) + geom_point() + facet_wrap( ~ id)
```
 
  1. Select the id 12 and fit a Poisson distribution to the number of seizures for this patient.

  2. Implement and use a EM algorithm for fitting a mixture of 2 Poisson distributions to this data
  
  3. Compare the two models
  
